---
title: "Task 4 - Statistical Inference"
author: "IT3081 Statistical Modelling"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    df_print: paged
  html_notebook:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
subtitle: Retail Product Demand Analysis at Fragceylon
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 6,
  fig.align = "center"
)

options(scipen = 999)

project_root = "C:/Users/saths/Desktop/Sliit/3rd year/3rd Year 1st sem/IT3081-SM/Project/IT3081-Fragceylon-Retail-Demand-Analysis"
knitr::opts_knit$set(root.dir = project_root)

library(readxl)
library(knitr)
library(car)
```

# 1. Purpose of Task 4

Task 4 uses statistical inference to determine whether selected differences observed in the descriptive analysis are large enough to be statistically distinguishable from random sample variation.

For all main tests:

\[
\alpha = 0.05
\]

The response variable is **Regular Order Quantity**.

# 2. Load the Analytical Dataset

```{r load-data}
regular_file = "04_Cleaned_Data/Fragceylon_Regular_Orders.xlsx"

regular_orders = read_excel(regular_file)
regular_orders$Date = as.Date(regular_orders$Date)

dim(regular_orders)
summary(regular_orders$Quantity)
```

# 3. Test 1 - Do Mean Order Quantities Differ by Product?

## 3.1 Business question

Do the four perfumes have different average Regular Order quantities?

## 3.2 Why this test?

The response variable, Quantity, is numerical and Product Name contains four independent categories. A one-way ANOVA is therefore appropriate for comparing the group means.

## 3.3 Hypotheses

**H0:** Mean Regular Order Quantity is equal across all four products.

**H1:** At least one product has a different mean Regular Order Quantity.

## 3.4 Group summaries

```{r product-summary}
product_group_summary = aggregate(
  Quantity ~ `Product Name`,
  data = regular_orders,
  FUN = function(x) c(
    N = length(x),
    Mean = mean(x),
    SD = sd(x)
  )
)

product_group_summary
```

## 3.5 ANOVA

```{r product-anova}
regular_orders$`Product Name` = factor(
  regular_orders$`Product Name`
)

product_anova = aov(
  Quantity ~ `Product Name`,
  data = regular_orders
)

summary(product_anova)
```

## 3.6 Effect size

```{r product-eta}
anova_table = summary(product_anova)[[1]]

ss_between = anova_table["`Product Name`", "Sum Sq"]
ss_total = sum(anova_table[, "Sum Sq"])

eta_squared = ss_between / ss_total
eta_squared
```

## 3.7 Result and interpretation

The project result is approximately:

- **F(3, 3112) = 0.3821**
- **p = 0.7659**
- **eta-squared ≈ 0.00037**

Because p > 0.05, we **fail to reject H0**. There is no statistically significant evidence that mean transaction-level Regular Order Quantity differs among the four perfumes.

The extremely small effect size also indicates that Product Name explains very little of the overall transaction-level Quantity variation.

# 4. Test 2 - Do Product Quantity Variances Differ?

## 4.1 Business question

Even if average order sizes are similar, do some products have more variable order quantities than others?

## 4.2 Why Brown-Forsythe?

The Brown-Forsythe version of Levene's test is robust to non-normality because it uses the group median as the centre.

## 4.3 Hypotheses

**H0:** Quantity variance is equal across all four products.

**H1:** At least one product has a different Quantity variance.

```{r product-bf}
product_bf = leveneTest(
  Quantity ~ `Product Name`,
  data = regular_orders,
  center = median
)

product_bf
```

## 4.4 Result and interpretation

The project result is approximately:

- **F = 0.1349**
- **p = 0.9393**

Because p > 0.05, we **fail to reject H0**. The data do not provide evidence that Quantity variability differs by product.

# 5. Test 3 - Controlled Local vs Japan Mean Quantity

## 5.1 Business question

Within a more comparable operating context, are Japan-labelled transactions associated with different average order quantities than Local transactions?

## 5.2 Why restrict the data?

Japan records occur only after 1 April 2025, within the Main Dealer structure, and are concentrated in one buyer. To reduce structural confounding, the comparison is restricted to:

- Main Dealer
- Post-April-2025 regime
- Buyer = Prabath Liyanarachchi

This still does **not** establish causality. It is a controlled observational comparison.

## 5.3 Why Welch's t-test?

The response is numerical and the comparison contains two independent groups. Welch's t-test is preferred because it does not require equal population variances.

## 5.4 Hypotheses

**H0:** Mean Quantity is equal between Japan and Local transactions.

**H1:** Mean Quantity differs between Japan and Local transactions.

```{r market-subset}
market_df = regular_orders[
  regular_orders$Buyer == "Prabath Liyanarachchi" &
  regular_orders$`Channel Type` == "Main Dealer" &
  regular_orders$Sales_Regime == "Post-April-2025",
]

table(market_df$`Export Market`)

aggregate(
  Quantity ~ `Export Market`,
  data = market_df,
  FUN = mean
)
```

```{r market-welch}
market_test = t.test(
  Quantity ~ `Export Market`,
  data = market_df,
  var.equal = FALSE,
  conf.level = 0.95
)

market_test
```

## 5.5 Effect size - Hedges' g

```{r market-effect}
japan = market_df$Quantity[
  market_df$`Export Market` == "Japan"
]

local = market_df$Quantity[
  market_df$`Export Market` == "Local"
]

n1 = length(japan)
n2 = length(local)

sp = sqrt(
  ((n1 - 1) * var(japan) + (n2 - 1) * var(local)) /
  (n1 + n2 - 2)
)

cohen_d = (mean(japan) - mean(local)) / sp

J = 1 - 3 / (4 * (n1 + n2) - 9)
hedges_g = J * cohen_d

hedges_g
```

## 5.6 Result and interpretation

The project result is approximately:

- Japan mean = **622.53**
- Local mean = **283.70**
- Difference = **338.83**
- **t(308.82) = 18.0612**
- **p < 0.001**
- 95% CI for the mean difference ≈ **301.91 to 375.74**
- Hedges' g ≈ **1.8845**

We **reject H0**. Japan-labelled transactions in this controlled subset have substantially larger average transaction Quantity than Local transactions.

**Important:** This is an association. The Japan indicator is highly concentrated in one buyer and a particular business structure, so the result must not be described as a causal effect of export market.

# 6. Test 4 - Pre-April Shop vs Small Dealer Mean Quantity

## 6.1 Business question

Before the April 2025 structural change, did Shop and Small Dealer transactions differ in average order size?

## 6.2 Why restrict to the pre-April period?

After 1 April 2025, the dataset contains only Main Dealer transactions. Restricting this comparison to the earlier regime avoids mixing fundamentally different channel structures.

## 6.3 Why Welch's t-test?

There are two independent channel groups with numerical Quantity, and equal variance is not assumed.

## 6.4 Hypotheses

**H0:** Mean Quantity is equal for Shop and Small Dealer transactions.

**H1:** Mean Quantity differs between Shop and Small Dealer transactions.

```{r channel-subset}
channel_df = regular_orders[
  regular_orders$Sales_Regime == "Pre-April-2025" &
  regular_orders$`Channel Type` %in% c("Shop", "Small Dealer"),
]

table(channel_df$`Channel Type`)

aggregate(
  Quantity ~ `Channel Type`,
  data = channel_df,
  FUN = mean
)
```

```{r channel-welch}
channel_test = t.test(
  Quantity ~ `Channel Type`,
  data = channel_df,
  var.equal = FALSE,
  conf.level = 0.95
)

channel_test
```

## 6.5 Result and interpretation

The project result is approximately:

- Small Dealer mean = **173.12**
- Shop mean = **29.89**
- Difference = **143.23**
- **t(351.14) = 37.0431**
- **p < 0.001**
- 95% CI ≈ **135.63 to 150.83**
- Hedges' g ≈ **3.4927**

We **reject H0**. During the pre-April-2025 regime, Small Dealer transactions had much larger average Quantity than Shop transactions.

# 7. Test 5 - Does Post-April Quantity Differ by Calendar Month?

## 7.1 Business question

Within the current Main Dealer regime, is there evidence of systematic differences in mean transaction Quantity across calendar months?

## 7.2 Why Welch one-way ANOVA?

There are more than two calendar-month groups and the response is numerical. Welch's one-way ANOVA is used because it does not require equal group variances.

## 7.3 Hypotheses

**H0:** Mean Quantity is equal across represented calendar months.

**H1:** At least one calendar month has a different mean Quantity.

```{r post-month-data}
post_df = regular_orders[
  regular_orders$Sales_Regime == "Post-April-2025",
]

post_df$Month = factor(
  post_df$Month,
  levels = month.name
)

table(post_df$Month)
```

```{r month-welch-anova}
month_test = oneway.test(
  Quantity ~ Month,
  data = post_df,
  var.equal = FALSE
)

month_test
```

## 7.4 Result and interpretation

The project result is approximately:

- **F(11, 173.55) = 1.1124**
- **p = 0.3542**

Because p > 0.05, we **fail to reject H0**. There is no statistically significant evidence that average transaction Quantity differs across calendar months in the post-April Main Dealer regime.

# 8. Supporting Robustness Checks

Robustness checks are useful because a statistical conclusion should not depend entirely on one narrow modelling choice.

## 8.1 Product ANOVA after removing IQR high outliers

```{r robustness-product-outliers}
product_no_outliers = regular_orders[
  regular_orders$Outlier_IQR_Flag != "Yes",
]

summary(
  aov(
    Quantity ~ `Product Name`,
    data = product_no_outliers
  )
)
```

## 8.2 Product ANOVA within the post-April regime

```{r robustness-product-post}
summary(
  aov(
    Quantity ~ `Product Name`,
    data = post_df
  )
)
```

## 8.3 Local vs Japan after excluding IQR high outliers

```{r robustness-market-outliers}
market_no_outliers = market_df[
  market_df$Outlier_IQR_Flag != "Yes",
]

t.test(
  Quantity ~ `Export Market`,
  data = market_no_outliers,
  var.equal = FALSE
)
```

# 9. Final Statistical Inference Summary

```{r inference-summary}
inference_summary = data.frame(
  Test = c(
    "Product mean Quantity - One-Way ANOVA",
    "Product variance - Brown-Forsythe",
    "Controlled Local vs Japan - Welch t-test",
    "Pre-April Shop vs Small Dealer - Welch t-test",
    "Post-April calendar month - Welch ANOVA"
  ),
  Approx_p_value = c(
    0.7659,
    0.9393,
    "<0.001",
    "<0.001",
    0.3542
  ),
  Decision = c(
    "Fail to reject H0",
    "Fail to reject H0",
    "Reject H0",
    "Reject H0",
    "Fail to reject H0"
  )
)

kable(
  inference_summary,
  caption = "Task 4 Main Inferential Findings"
)
```

# 10. Task 4 Main Conclusion

The strongest inferential differences are associated with **transaction structure** rather than Product Name or calendar month. Product means and product variances are statistically similar, while substantial differences appear between Local and Japan-labelled transactions in the controlled subset and between Shop and Small Dealer transactions in the pre-April regime.

These findings are based on historical observational data. They identify associations, not causal effects.
